---
title: "Reproducible Research: Peer Assessment 1"
author: "Fitz Gerald M. Villegas"
date: "9/27/2020"
output: md_document
---
1.Loading and Preprocessing the data
```{r echo = TRUE}
#Download data from Web
download.file("https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2Factivity.zip", destfile = "activity.zip", mode="wb")
#Unzip Data and and store 
unzip("activity.zip")
dataset <- read.csv("activity.csv", header = TRUE)
head(dataset)
```
-> Process/transform the data (if necessary) into a format suitable for your analysis
```{r echo = TRUE}
setDate <- transform(dataset, date = as.Date(date))
str(setDate)
```
2. What is mean total number of steps taken per day?

->Ignoring missing values in the dataset.
```{r echo = TRUE}
blank <- is.na(dataset$steps)
setDate$isNA <- blank
newdata <- setDate[setDate$isNA != 'TRUE',]
nrow(newdata)
```

->Calculating the total number of steps taken per day and creating a histogram.
```{r echo = TRUE}
dataperday<-aggregate(steps~date,newdata,sum, na.rm = TRUE)
hist(dataperday$steps, main="Steps Taken per Day", xlab="Total Steps per Day", ylab = "Number of Days", col = "gray")
```

->Calculating the Mean and Median Values
```{r echo=TRUE}
mean <- mean(dataperday$steps)
cat("Mean for total steps taken per day: ", mean, fill=TRUE)
median <- median(dataperday$steps)
cat("Median for total steps taken per day: ", median, fill=TRUE)
```

3. What is the average daily activity pattern?

-> Making a time series plot (i.e. type = "l") of the 5-minute interval (x-axis) and the average number of steps taken, averaged across all days (y-axis)
```{r echo = TRUE}
avgsteps<-aggregate(steps~interval,newdata,sum, na.rm = TRUE)
plot(x = avgsteps$interval,y = avgsteps$steps,type = "l",xlab = "5 Minute Interval", ylab = "Average Number of Steps",main = "Average Daily Activity Planner")
```

->Getting the average maximum steps per day in 5-minute interval
```{r echo = TRUE}
max_avgsteps <- avgsteps$interval[avgsteps$steps == max(avgsteps$steps)]
cat("5-min time Interval with maximum average steps taken per day: ", max_avgsteps, fill = TRUE)
```

4. Imputing missing values

->Calculating the total number of missing values in the dataset
```{r echo=TRUE}
blank_count <- sum(!complete.cases(dataset))
cat("Missing value count for rows: ", blank_count, fill = TRUE)
```

->Filling in the missing date using mean data for missing values
```{r echo=TRUE}
tofilldataset <- dataset
sapply(unique(dataset$interval), function(x) tofilldataset[!complete.cases(tofilldataset) & (tofilldataset$interval == x),1] <- avgsteps$steps[avgsteps$interval == x])
```

->Calculating the total number of steps taken per day and creating a histogram.
```{r echo = TRUE}
secondsteps <- aggregate(steps~date,tofilldataset,sum)
hist(secondsteps$steps,main = "Steps per Day", xlab = "Total Steps per Day", ylab = "Number of Days", col = "gray")
```

->Calculating the Mean and Median Values and Comparing to the previous Mean and Median
```{r echo = TRUE}
newmean <- mean(secondsteps$steps)
cat("Old mean is: ", mean, fill = TRUE)
cat("Mean total steps taken per day: ", newmean, fill = TRUE)

newmedian <- median(secondsteps$steps)
cat("Old median is: ", median, fill = TRUE)
cat("Median total steps taken per day: ", newmedian, fill = TRUE)
```

5. Are there differences in activity patterns between weekdays and weekends?
```{r echo = TRUE}
setDate$weekend <- "weekday"
setDate$weekend[weekdays(setDate$date) %in% c("Saturday", "Sunday")] <- "weekend"
setDate$weekend <- as.factor(setDate$weekend)
thirdstep <- aggregate(steps~interval + weekend,data = setDate,sum, na.rm = TRUE)
```

->Making a panel plot containing a time series plot (i.e. type = "l") of the 5-minute interval (x-axis) and the average number of steps taken, averaged across all weekday days or weekend days (y-axis). 
```{r echo = TRUE}
library(lattice)
xyplot(steps~interval | weekend,thirdstep,type = "l", layout = c(1,2),xlab = "Interval", ylab = "Number of Steps")
```

